---
title: "Characteristics of Distributions"
subtitle: "DKU Stats 101 Spring 2022"
author: "Professor MacDonald"
date: "1/11/2022"
output: 
  learnr::tutorial:
    toc_depth: 2
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

library(tidyverse)
library(gapminder)
library(learnr)

theme_set(theme_classic())

modes <- function(x) {
  ux <- unique(x)
  tab <- tabulate(match(x, ux))
  ux[tab == max(tab)]
}
```

# Characteristics of distributions

## Distribution of common quantities

Many phenomena in nature have a relatively easily guessed distribution characteristics

- What is the distribution of length of rivers in the U.S.?
- What is the distribution of width of flower sepals? 
- What is the distribution of life expectancy across countries in 2007?

Features to guess: 

- Shape
- Center
- Spread

```{r importroster}
classroster <- read.csv("www/classroster.csv", fileEncoding="UTF-8-BOM")
```


```{r picker1, exercise=TRUE}
sample(classroster$name, 1)
```

## Graphs of common quantities

### Length of rivers in the U.S.

```{r river_shape, exercise=TRUE}
ggplot(data.frame(rivers), aes(x=rivers)) + 
  geom_histogram(fill="blue4", color="black") +
  labs(x="Length in miles", y="count")
```

### Flower sepal width

```{r flower_width, exercise=TRUE}
ggplot(iris, aes(x=Sepal.Width)) + 
  geom_histogram(fill="blue4", color="black", bins=15) +
  labs(x="Sepal width", y="Count")
```

### Life expetancy in 2007

```{r life_exp, exercise=TRUE}
gap2007 <-gapminder %>%
  filter(year==2007)

ggplot(gap2007, aes(x=lifeExp)) + 
  geom_histogram(fill="blue4", color="black") +
  labs(x="Life expectancy in years", y="Count")
```

## Distribution of our data

Let's now collect some data about our class

- Height
- Information about handedness

## Guessing the shape of our data

Take a guess at what each question's distribution characteristics will be:

- Shape
  + Skew 
  + Modes
- Center
  + Mean
  + Median
- Spread
  + Range
  + IQR
  + Standard deviation

- Also think carefully about the difference between the three different calculations of handedness - how do they differ? Discuss with your partner.

## Height summary statistics

```{r importdata}
ourdata <- read.csv("www/ourdata.csv")
```

- Shape
  + Skew:  
  + Modes: **`r modes(ourdata$Q1)`** 
- Center
  + Mean **`r mean(ourdata$Q1, na.rm=TRUE)`**
  + Median **`r median(ourdata$Q1, na.rm=TRUE)`**
- Spread
  + Range **`r range(ourdata$Q1, na.rm=TRUE)`**
  + IQR **`r IQR(ourdata$Q1, na.rm=TRUE)`**
  + Standard deviation **`r sd(ourdata$Q1, na.rm=TRUE)`**

```{r picker2, exercise=TRUE}
sample(classroster$name, 1)
```

## Height graph

```{r heights, exercise=TRUE}
ggplot(ourdata, aes(x=Q1)) +
  geom_histogram(fill="blue4", color="black", bins=6) + 
  labs(x="Height in cm", y="Count")
```

## Handedness l-r summary statistics

- Shape
  + Skew: 
  + Modes: **`r modes(ourdata$Q2)`** 
- Center
  + Mean **`r mean(ourdata$Q2)`**
  + Median **`r median(ourdata$Q2, na.rm=TRUE)`**
- Spread
  + Range **`r range(ourdata$Q2, na.rm=TRUE)`**
  + IQR **`r IQR(ourdata$Q2, na.rm=TRUE)`**
  + Standard deviation **`r sd(ourdata$Q2, na.rm=TRUE)`**

- What does this measure? 

```{r picker3, exercise=TRUE}
sample(classroster$name, 1)
```

## Handedness l-r graph

```{r lminusr, exercise=TRUE}
ggplot(ourdata, aes(x=Q2)) +
  geom_histogram(fill="blue4", color="black", binwidth=2) + 
  labs(x="Handedness left - right", y="Count")
```

## Handedness l+r summary statistics

- Shape
  + Skew: 
  + Modes: **`r modes(ourdata$Q3)`** 
- Center
  + Mean **`r mean(ourdata$Q3, na.rm=TRUE)`**
  + Median **`r median(ourdata$Q3, na.rm=TRUE)`**
- Spread
  + Range **`r range(ourdata$Q3, na.rm=TRUE)`**
  + IQR **`r IQR(ourdata$Q3, na.rm=TRUE)`**
  + Standard deviation **`r sd(ourdata$Q3, na.rm=TRUE)`**

- What does this measure? 

```{r picker4, exercise=TRUE}
sample(classroster$name, 1)
```

## Handedness l+r graph

```{r lplusr, exercise=TRUE}
ggplot(ourdata, aes(x=Q3)) +
  geom_histogram(fill="blue4", color="black", binwidth=2) + 
  labs(x="Handedness left + right", y="Count")
```

### Handedness left - right / left + right graph

```{r lrratio, exercise=TRUE}
ggplot(ourdata, aes(x=Q4)) +
  geom_histogram(fill="blue4", color="black", binwidth=2) + 
  labs(x="Handedness left - right / left + right", y="Count")
```

## Closing thoughts

- Many distributions can be guessed in advanced based on the data generating process
- You should have at least a guess as to what the distribution is before starting your exploratory data analysis
- Think carefully about what your variable is actually measuring
- Characteristics of distributions are **summaries** of the data, almost always obscure features of the data
  + Don't mislead your readers!!